---
title: "State-of-Life-Predicting-Life-Expectancy-from-1970s-Census-Data"
author: "XunSun"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
library(tidyverse)
library(GGally)
library(patchwork)
library(gt)
library(leaps)
library(caret)
```
R dataset state.x77 from library(faraway) contains information on 50 states from 1970s collected by US Census Bureau. The goal is to predict ‘life expectancy’ using a combination of remaining variables.
```{r}
library(faraway)
data.state <- as.data.frame(state.x77)|> 
  janitor::clean_names() 
head(data.state)
view(data.state)
```
a) Provide descriptive statistics for all variables of interest – no test required
```{r}
summary(data.state)
```
Examine exploratory plots, e.g., scatter plots, histograms, boxplots to get a sense of
the data and possible variable transformations.
```{r}
data.state |>
  relocate(`life_exp`) |>
  ggpairs()
```
## look for appropriate transformations
```{r}
density_plot_population <-data.state |>
  ggplot(aes(x = population)) +
  geom_density()

density_plot_logpopulation <-dat.state |>
  mutate(log_population = log(population)) |>
  ggplot(aes(x = log_population)) + geom_density()

density_plot_inver_population <-dat.state |>
  mutate(inver_population = 1/(population)) |>
  ggplot(aes(x = inver_population)) + geom_density()


density_plot_population + density_plot_logpopulation + density_plot_inver_population
```

```{r}
density_plot_area <- data.state |>
  ggplot(aes(x = area)) +
  geom_density()

density_plot_logarea <- dat.state |>
  mutate(log_area = log(area)) |>
  ggplot(aes(x = log_area)) + geom_density()

density_plot_area + density_plot_logarea
```

```{r}
density_plot_illiteracy <- data.state |>
  ggplot(aes(x = illiteracy)) +
  geom_density()

density_plot_logilliteracy <- dat.state |>
  mutate(log_illiteracy = log(illiteracy)) |>
  ggplot(aes(x = log_illiteracy)) + geom_density()

density_plot_illiteracy + density_plot_logilliteracy
```


```{r}
qq_population <- data.state |>
  ggplot(aes(sample = population)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Population")

qq_log_population <- data.state |>
  mutate(log_population = log(population)) |>
  ggplot(aes(sample = log_population)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Log Population")

qq_inver_population <- data.state |>
  mutate(inver_population = 1 / (population)) |>
  ggplot(aes(sample = inver_population)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Inverse Population")

# QQ Plots for Area
qq_area <- data.state |>
  ggplot(aes(sample = area)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Area")

qq_log_area <- data.state |>
  mutate(log_area = log(area)) |>
  ggplot(aes(sample = log_area)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Log Area")

# QQ Plots for Illiteracy
qq_illiteracy <- data.state |>
  ggplot(aes(sample = illiteracy)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Illiteracy")

qq_log_illiteracy <- data.state |>
  mutate(log_illiteracy = log(illiteracy)) |>
  ggplot(aes(sample = log_illiteracy)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Plot: Log Illiteracy")

# Combine QQ Plots with Patchwork for Comparison
library(patchwork)

(qq_population | qq_log_population | qq_inver_population) / 
(qq_area | qq_log_area) / 
(qq_illiteracy | qq_log_illiteracy)
```

```{r}
data.state <- data.state |>
  mutate(log_Population = log(population)) 
view(data.state)
```


```{r}
# Define the response variable and dataset
response <- "life_expectancy"  # Replace with your actual response variable
data <- data.state            # Your dataset

# Null model
null_model <- lm(life_exp ~ 1, data = data)

# Full model (all predictors)
full_model <- lm(life_exp ~ ., data = data)

```

```{r}

library(MASS)
library(olsrr)

```
```{r}
# Forward Selection
forward_model_aic <- stepAIC(null_model, 
                         scope = list(lower = null_model, upper = full_model), direction = "forward")

# Summary of the selected model
summary(forward_model_aic)


```
Backward Selection
```{r}
# Backward selection based on AIC
backward_aic <- step(full_model, 
                     direction = "backward",trace = TRUE)

# Display summary of the final model
summary(backward_aic)

# Number of observations
n <- nrow(data.state)

# Backward selection based on BIC
backward_bic <- step(full_model, 
                     direction = "backward", 
                     k = log(n), 
                     trace = TRUE)

# Display summary of the final model
summary(backward_bic)


# Forward Selection with p-values
forward_model_ols <- ols_step_forward_p(null_model, 
                                        full_model, 
                                        p_enter = 0.05)

# Summary and results
print(forward_model_ols)

```

```{r}
# Stepwise selection (both directions, default AIC)
stepwise_aic <- step(full_model, 
                     direction = "both",trace = TRUE)

# Summary of the final model
summary(stepwise_aic)

# Number of observations
n <- nrow(data.state)

# Stepwise selection (both directions, BIC)
stepwise_bic <- step(full_model, 
                     direction = "both", 
                     k = log(n),trace = TRUE)  # BIC penalty

# Summary of the final model
summary(stepwise_bic)

```

```{r}
library(olsrr)

# Perform stepwise selection based on p-values (default uses AIC)
stepwise_model <- ols_step_both_p(full_model, 
                                  pent = 0.05,  # Entry significance level
                                  prem = 0.05) # Removal significance level

# Print stepwise selection results
print(stepwise_model)
```

